---
title: "Disaster Declaration Analysis"
author: "Vincent"
date: "2023-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries used}
library(tidyverse)
library(RSQLite)
library(DBI)
library(ggplot2)
library(dplyr)
```

###Connect to DB and set working directory
##build a data frame from the queries
```{r set working directory}
setwd("~/workbook")

con <- dbConnect(RSQLite::SQLite(),"Disaster_Data.db")

dbListTables(con)
dbListFields(con,"us_disaster_declarations")

res<- dbSendQuery(con,
"SELECT fy_declared, 
count(incident_type) AS number_of_incidents 
FROM us_disaster_declarations
WHERE fy_declared > 1960
GROUP BY fy_declared
ORDER by fy_declared DESC;")

res<- dbGetQuery(con,
"SELECT fy_declared, 
count(incident_type) AS number_of_incidents 
FROM us_disaster_declarations
WHERE fy_declared > 1960
GROUP BY fy_declared
ORDER by fy_declared DESC;")
```

#build data frames and summary statics
```{r build the dataframes}
disaster_df <- dbReadTable(con, "us_disaster_declarations")
type<- disaster_df %>% 
  count(incident_type)

summary(disaster_df)
colnames(disaster_df)
head(disaster_df)
View(disaster_df)
```

###rename and rearrange columns
```{r rename columns}
res <- res %>% 
  rename(Number_of_Disasters_Declared = number_of_incidents,Year_Declared = fy_declared)

type <- type %>% 
  arrange(n)
```


#take a look at the distribution of the graphs this this test for outliers
```{r search for outliers }
boxplot(res$Number_of_Disasters_Declared)
boxplot(type$n)
```


#filter out outliers
#outliers filters where the year 2020(for covid-19), and 2023(because it was incomplete collection of information) 
```{r filters}
res <- res %>% 
  filter(Number_of_Disasters_Declared < 4500,Year_Declared < 2023)
```


#see changed distrubtion
```{r check outliers again}
boxplot(res$Number_of_Disasters_Declared)
```

#build an arrangment paired with mutaute in DPLYR library than pass it to ggplot and build the graph
#this reorders the factors in the graph so its easier to break down and look at
```{r build graph}
type %>% 
  arrange(n) %>% 
  mutate(incident_type=factor(incident_type,levels=incident_type)) %>% 
  ggplot(aes(x=incident_type,y=n))+
  geom_point(aes())+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#build a frequncy table to see the distrubtion.
```{build a frequency table}
hist(res$Number_of_Disasters_Declared)
```


#build a standard GGPLOT since this graph is on a timelines no need to do anything to it
```{r build out the dotplot}
res.graph <-ggplot(data = res,mapping = aes(x = Year_Declared,y=Number_of_Disasters_Declared))+
  geom_point(aes())+
  geom_smooth(method = "lm",col="black")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(title = "Number of Natural Disasters Declared by year",subtitle ="By the federal government")
```

```{r}
res.graph

cor(res$Year_Declared,res$Number_of_Disasters_Declared)

cor.test(res$Year_Declared,res$Number_of_Disasters_Declared)
```


```{r disconnect from the DB}
dbDisconnect(con)
```


