---
title: "Disaster Declaration Analysis"
author: "Vincent"
date: "2023-02-01"
output: html_document
---


```{r libraries used, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(RSQLite)
library(DBI)
library(ggplot2)
library(dplyr)
```

### Connect to DB and set working directory
### build a data frame from the queries
```{r set working directory, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
setwd("~/workbook")

con <- dbConnect(RSQLite::SQLite(),"Disaster_Data.db")

dbListTables(con)
dbListFields(con,"us_disaster_declarations")

res<- dbSendQuery(con,
"SELECT fy_declared, 
count(incident_type) AS number_of_incidents 
FROM us_disaster_declarations
WHERE fy_declared > 1960
GROUP BY fy_declared
ORDER by fy_declared DESC;")

res<- dbGetQuery(con,
"SELECT fy_declared, 
count(incident_type) AS number_of_incidents 
FROM us_disaster_declarations
WHERE fy_declared > 1960
GROUP BY fy_declared
ORDER by fy_declared DESC;")
```

### build data frames and summary statics

```{r build the dataframes, echo=FALSE, message=FALSE, warning=FALSE}
disaster_df <- as.tibble(dbReadTable(con, "us_disaster_declarations"))
type<- disaster_df %>% 
  count(incident_type)

head(disaster_df)

# Summary for Numeric Columns
numeric_summary <- disaster_df %>%
  summarise(across(where(is.numeric), list(
    mean = ~mean(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE),
    min = ~min(., na.rm = TRUE),
    max = ~max(., na.rm = TRUE),
    n_missing = ~sum(is.na(.))
  )))

# Summary for Character Columns
character_summary <- disaster_df %>%
  summarise(across(where(is.character), list(
    unique_count = ~n_distinct(.),
    most_common = ~names(sort(table(.), decreasing = TRUE)[1]),
    n_missing = ~sum(is.na(.))
  )))

# Print summaries
print(numeric_summary)
print(character_summary)

```

### rename and rearrange columns
```{r rename columns}
res <- res %>% 
  rename(Number_of_Disasters_Declared = number_of_incidents,Year_Declared = fy_declared)

type <- type %>% 
  arrange(n)
```


### take a look at the distribution of the graphs this this test for outliers
```{r search for outliers, echo=FALSE, message=FALSE, warning=FALSE}
boxplot(res$Number_of_Disasters_Declared)
boxplot(type$n)
```


### filter out outliers
### outliers filters where the year 2020(for covid-19), and 2023(because it was incomplete collection of information) 
```{r filters, include=FALSE}
res <- res %>% 
  filter(Number_of_Disasters_Declared < 4500,Year_Declared < 2023)
```


### see changed distrubtion
```{r check outliers again, echo=FALSE, message=FALSE, warning=FALSE}
boxplot(res$Number_of_Disasters_Declared)
```

### build an arrangment paired with mutaute in DPLYR library than pass it to ggplot and build the graph
### this reorders the factors in the graph so its easier to break down and look at
```{r build graph, echo=FALSE, message=FALSE, warning=FALSE}
# Assuming disaster_df is your dataframe
# Reordering the factor based on count
disaster_df <- disaster_df %>%
  mutate(incident_type = reorder(incident_type, incident_type, FUN = function(x) -length(x)))

# Plotting
ggplot(disaster_df, aes(x = incident_type)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Frequency of Different Types of Disasters",
       x = "Disaster Type",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# build a frequncy table to see the distrubtion.
```{build a frequency table}
hist(res$Number_of_Disasters_Declared)
```


### build a standard GGPLOT since this graph is on a timelines no need to do anything to it
```{r build out the dotplot, echo=FALSE, message=FALSE, warning=FALSE}
res.graph <-ggplot(data = res,mapping = aes(x = Year_Declared,y=Number_of_Disasters_Declared))+
  geom_point(aes())+
  geom_smooth(method = "lm",col="black")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(title = "Number of Natural Disasters Declared by year",subtitle ="By the federal government")

print(res.graph)
```
```{r Geographic Analysis}

ggplot(disaster_df, aes(x = state, fill = incident_type)) +
  geom_bar() +
  labs(title = "Distribution of Disaster Types by State",
       x = "State",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```



```{r regression model, echo=FALSE, warning=FALSE}

# Linear regression analysis
model <- lm(Number_of_Disasters_Declared ~ Year_Declared, data = res)

# Summary of the model
summary(model)


ggplot(res, aes(x = Year_Declared, y = Number_of_Disasters_Declared)) +
  geom_point() +  # Scatter plot points
  geom_smooth(method = "lm", col = "blue") +  # Regression line
  labs(title = "Regression Analysis",
       x = "Year Declared",
       y = "Number of Disasters Declared") +
  theme_minimal()

```
```{r residuals vs fitted model}

# Assuming 'model' is your linear model
ggplot(data = res, aes(x = fitted(model), y = resid(model))) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()


```


```{r disconnect from the DB, include=FALSE}
dbDisconnect(con)
```


